---
title: "eds223-week5-section"
format: html
---

# Practice raster operations with vectors

### Set up 
```{r}
# Load libraries 
library(tidyverse)
library(sf)
library(terra)
library(spData)
library(spDataLarge)
library(tmap)

# Load raster data representing grain sizes with the three classes clay, silt and sand
grain <- terra::rast(system.file("raster/grain.tif", package = "spData"))
nz_height <- spData::nz_height
```



```{r}
class(grain) # terra raster 
class(nz_height) # sf data frame 

st_crs(nz_height)
```

Subset points higher than 3100 meters in nz_height (from spData) and create a template raster with rast(), where the resolution is 3 km x 3 km for the extent of the subset dataset

- Count numbers of the highest points in each grid cell. 
- Find the maximum elevation in each grid cell 

```{r}
# Subset for elevation > 3100 
nz_height_subset <- nz_height %>% 
  filter(elevation > 3100)

# Create template raster 
nz_raster_template <- rast(terra::ext(nz_height_subset),
                      resolution = 3000,
                      crs = terra::crs(nz_height_subset))
```

```{r}
# Create raster that counts elevation points 
nz_raster <- rasterize(nz_height_subset, nz_raster_template,
                       field= "elevation",
                       fun= "count")

plot(nz_raster)
```

```{r}
# Find max elevation in each grid 
nz_raster2 <- rasterize(nz_height_subset, nz_raster_template,
                       field= "elevation",
                       fun= "max")

plot(nz_raster2)
plot(st_geometry(nz_height_subset), add = TRUE)
```

With the previous raster, complete the following:

- Aggregate the raster that counts the highest points in New Zealand/Aotearoa
- Reduce its geographic resolution by half, such that cells are 6 x 6 km
- Plot the result
- Resample back to the original resolution of 3 km x 3 km

```{r}
# Reduce the resolution by combining 2 cells in each direction into larger cells
# Sum the values of all cells for the resulting elevation value
nz_raster_low <- aggregate(nz_raster, fact = 2, fun = sum, na.rm = TRUE)

# Convert the new raster's resolution back to the 3kmx3km resolution of original raster
nz_resample <- resample(nz_raster_low, nz_raster)

plots <- c(nz_raster, nz_resample)
labs <- c("Original 6 x 6 km", "Resample 6 x 6 km")
plot(plots, main = labs)

plot(nz_raster_low, main = "Resample 3 x 3 km")
```

# Vectorize Raster 
```{r}
# Convert raster data to polygon vector data
grain_poly <- as.polygons(grain) %>% 
  st_as_sf()

plot(grain, main = "Grain (Raster)")
plot(grain_poly, main = "Grain (Vector)")
```

